---
title: "Advanced Statistical Analysis with descriptR"
author: "Abdul Ali"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Statistical Analysis with descriptR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette covers advanced statistical methods available in **descriptR**, including effect sizes, normality testing, missing data analysis, outlier detection, multivariate methods, and advanced modeling techniques.

```{r load-package, eval = FALSE}
library(descriptR)
library(dplyr) # For data manipulation
```

## Effect Sizes

Effect sizes quantify the magnitude of differences or relationships, providing context beyond statistical significance.

### Cohen's d (Independent Groups)

Measures the standardized mean difference between two groups:

```{r cohens-d, eval = FALSE}
# Compare mpg between 4-cylinder and 8-cylinder cars
mtcars_subset <- mtcars %>% filter(cyl %in% c(4, 8))

result <- compute_cohens_d(
  x = mtcars_subset$mpg[mtcars_subset$cyl == 4],
  y = mtcars_subset$mpg[mtcars_subset$cyl == 8],
  pooled = TRUE,
  hedges = FALSE
)

print(result)
# Effect size interpretation:
# Small: |d| = 0.2
# Medium: |d| = 0.5
# Large: |d| = 0.8
```

### Hedges' g (Bias-Corrected)

Use Hedges' g for small sample sizes:

```{r hedges-g, eval = FALSE}
result <- compute_cohens_d(
  x = mtcars_subset$mpg[mtcars_subset$cyl == 4],
  y = mtcars_subset$mpg[mtcars_subset$cyl == 8],
  pooled = TRUE,
  hedges = TRUE # Apply bias correction
)

print(result)
```

### Paired Comparisons

For repeated measures or matched pairs:

```{r paired-d, eval = FALSE}
# Example: Before and after measurements
before <- c(120, 125, 130, 135, 128)
after <- c(115, 118, 125, 130, 122)

result <- compute_cohens_d(
  x = before,
  y = after,
  paired = TRUE
)

print(result)
```

### ANOVA Effect Sizes

#### Eta-Squared (η²)

Proportion of total variance explained:

```{r eta-squared, eval = FALSE}
# Fit ANOVA model
model <- aov(mpg ~ factor(cyl), data = mtcars)

# Compute eta-squared
result <- compute_eta_squared(model, type = "partial")

print(result)
# Interpretation:
# Small: η² = 0.01
# Medium: η² = 0.06
# Large: η² = 0.14
```

#### Omega-Squared (ω²)

Less biased estimate than eta-squared:

```{r omega-squared, eval = FALSE}
result <- compute_omega_squared(model)

print(result)
```

### Categorical Association

#### Cramér's V

Measures association between categorical variables:

```{r cramers-v, eval = FALSE}
# Create categorical variables
mtcars_cat <- mtcars %>%
  mutate(
    cyl_cat = factor(cyl),
    am_cat = factor(am, labels = c("Auto", "Manual"))
  )

result <- compute_cramers_v(
  mtcars_cat$cyl_cat,
  mtcars_cat$am_cat,
  bias_correct = TRUE
)

print(result)
# Interpretation:
# Small: V = 0.1
# Medium: V = 0.3
# Large: V = 0.5
```

#### Phi Coefficient (2x2 Tables)

For binary variables:

```{r phi, eval = FALSE}
result <- compute_phi(
  mtcars_cat$am_cat,
  factor(mtcars$vs)
)

print(result)
```

### Effect Size Conversion

Convert between different effect size metrics:

```{r convert-effect, eval = FALSE}
# Convert Cohen's d to correlation
r <- convert_effect_size(
  value = 0.8,
  from = "cohens_d",
  to = "correlation"
)

# Convert odds ratio to Cohen's d
d <- convert_effect_size(
  value = 2.5,
  from = "odds_ratio",
  to = "cohens_d"
)
```

### Comprehensive Effect Size Analysis

Compute all applicable effect sizes:

```{r all-effects, eval = FALSE}
result <- compute_all_effect_sizes(
  data = mtcars,
  outcome = "mpg",
  group = "cyl",
  type = "auto" # Auto-detect appropriate effect sizes
)

print(result)
```

## Normality Assessment

Comprehensive normality testing with multiple methods.

### Multiple Normality Tests

```{r normality-multi, eval = FALSE}
result <- assess_normality(
  mtcars,
  vars = c("mpg", "hp", "wt"),
  tests = c("shapiro", "ks", "anderson", "jarque", "dagostino"),
  suggest_transformations = TRUE
)

# View test results
print(result$test_results)

# View overall assessment
print(result$summary)

# Transformation suggestions
print(result$transformations)

# Narrative insights
cat(paste(result$insights, collapse = "\n"))
```

### Individual Tests

#### Shapiro-Wilk Test

Most powerful test for small to medium samples:

```{r shapiro, eval = FALSE}
result <- assess_normality(
  mtcars,
  vars = "mpg",
  tests = "shapiro"
)

print(result$test_results)
```

#### Kolmogorov-Smirnov Test

Better for larger samples:

```{r ks-test, eval = FALSE}
result <- assess_normality(
  mtcars,
  vars = "mpg",
  tests = "ks"
)
```

#### Anderson-Darling Test

More sensitive to deviations in the tails:

```{r anderson, eval = FALSE}
result <- assess_normality(
  mtcars,
  vars = "mpg",
  tests = "anderson"
)
```

### Transformation Recommendations

When data is non-normal, descriptR suggests appropriate transformations:

```{r transformations, eval = FALSE}
# Assess normality with transformation suggestions
result <- assess_normality(
  mtcars,
  vars = "hp",
  suggest_transformations = TRUE
)

# View suggestions
print(result$transformations)

# Apply suggested transformation
if (!is.null(result$transformations)) {
  transform <- result$transformations$suggested_transform[1]

  if (transform == "log") {
    mtcars$hp_log <- log(mtcars$hp)
  } else if (transform == "sqrt") {
    mtcars$hp_sqrt <- sqrt(mtcars$hp)
  }
}
```

## Missing Data Analysis

Comprehensive missing data assessment with pattern detection and MCAR testing.

### Complete Missing Data Analysis

```{r missing-complete, eval = FALSE}
# Create data with missing values
data_na <- mtcars
data_na$mpg[c(1, 5, 10)] <- NA
data_na$hp[c(2, 5, 15)] <- NA
data_na$wt[c(5, 20)] <- NA

result <- analyze_missing(
  data_na,
  test_mcar = TRUE,
  pattern_analysis = TRUE,
  correlation_analysis = TRUE
)

# Summary statistics
print(result$summary)

# Missing patterns
print(result$patterns)

# MCAR test (Little's test)
print(result$mcar_test)

# Correlations between missingness indicators
print(result$correlations)

# Recommendations
cat(paste(result$recommendations, collapse = "\n"))
```

### Missing Data Patterns

Identify common patterns of missingness:

```{r missing-patterns, eval = FALSE}
# Analyze patterns
result <- analyze_missing(
  data_na,
  pattern_analysis = TRUE
)

# View most common patterns
print(head(result$patterns, 10))

# Variables with highest missingness
print(result$summary[order(-result$summary$pct_missing), ])
```

### MCAR Testing

Test if data is Missing Completely At Random:

```{r mcar-test, eval = FALSE}
result <- analyze_missing(
  data_na,
  test_mcar = TRUE
)

if (result$mcar_test$p_value > 0.05) {
  message("Data appears to be MCAR (p > .05)")
} else {
  message("Data may not be MCAR - investigate patterns")
}
```

## Outlier Detection

Multiple methods for detecting outliers with consensus approach.

### Multiple Detection Methods

```{r outliers-multi, eval = FALSE}
result <- detect_outliers(
  mtcars,
  vars = c("mpg", "hp", "wt"),
  methods = c("zscore", "iqr", "mad", "grubbs", "dixon", "rosner"),
  consensus = TRUE,
  min_methods = 2
)

# Overall summary
print(result$summary)

# Detailed results by method
print(result$details)

# Consensus outliers (flagged by multiple methods)
print(result$consensus)

# Insights
cat(paste(result$insights, collapse = "\n"))
```

### Individual Detection Methods

#### Z-Score Method

Flags observations > 3 SD from mean:

```{r zscore-outliers, eval = FALSE}
result <- detect_outliers(
  mtcars,
  vars = "mpg",
  methods = "zscore",
  threshold = 3
)

print(result$details)
```

#### IQR Method

Uses interquartile range (1.5 * IQR rule):

```{r iqr-outliers, eval = FALSE}
result <- detect_outliers(
  mtcars,
  vars = "mpg",
  methods = "iqr",
  multiplier = 1.5
)
```

#### MAD Method

Median Absolute Deviation (robust to outliers):

```{r mad-outliers, eval = FALSE}
result <- detect_outliers(
  mtcars,
  vars = "mpg",
  methods = "mad"
)
```

#### Grubbs' Test

Statistical test for a single outlier:

```{r grubbs-outliers, eval = FALSE}
result <- detect_outliers(
  mtcars,
  vars = "mpg",
  methods = "grubbs"
)
```

### Consensus Approach

Require agreement from multiple methods:

```{r consensus-outliers, eval = FALSE}
# Require at least 3 methods to agree
result <- detect_outliers(
  mtcars,
  vars = "mpg",
  methods = c("zscore", "iqr", "mad", "grubbs"),
  consensus = TRUE,
  min_methods = 3
)

# Only show robust outliers
print(result$consensus)
```

## Principal Component Analysis

Dimensionality reduction with comprehensive interpretation.

### Basic PCA

```{r pca-basic, eval = FALSE}
result <- perform_pca(
  mtcars,
  vars = c("mpg", "hp", "wt", "disp", "drat"),
  scale = TRUE,
  n_components = NULL, # Keep all
  variance_threshold = 0.95
)

# Variance explained
print(result$variance_explained)

# Loadings
print(result$loadings)

# Insights
cat(paste(result$insights, collapse = "\n"))
```

### Component Selection

#### Kaiser Criterion

Keep components with eigenvalues > 1:

```{r kaiser, eval = FALSE}
result <- perform_pca(
  mtcars,
  vars = c("mpg", "hp", "wt", "disp", "drat"),
  n_components = "kaiser"
)

print(paste("Components retained:", result$n_components_retained))
```

#### Variance Threshold

Keep components explaining target variance:

```{r variance-threshold, eval = FALSE}
result <- perform_pca(
  mtcars,
  variance_threshold = 0.80
)

print(paste("Components for 80% variance:", result$n_components_retained))
```

### Component Interpretation

```{r pca-interpret, eval = FALSE}
result <- perform_pca(mtcars)

# Interpret first component
interp1 <- interpret_pca_component(result, component = 1, threshold = 0.3)
print(interp1)

# Get high loadings
loadings1 <- get_pca_loadings(result, component = 1, threshold = 0.4)
print(loadings1)
```

### PCA Scores

Extract component scores for further analysis:

```{r pca-scores, eval = FALSE}
result <- perform_pca(mtcars)

# Get all scores
scores <- get_pca_scores(result)

# Get specific components
scores_12 <- get_pca_scores(result, components = c(1, 2))

# Add to original data
mtcars_pca <- cbind(mtcars, scores_12)
```

### Rotation Methods

Apply rotation for interpretability:

```{r pca-rotation, eval = FALSE}
# Varimax rotation (orthogonal)
result_varimax <- perform_pca(
  mtcars,
  rotation = "varimax"
)

# Oblique rotation
result_promax <- perform_pca(
  mtcars,
  rotation = "promax"
)

# Compare loadings
print(result_varimax$loadings)
print(result_promax$loadings)
```

## Exploratory Factor Analysis

Extract latent factors from observed variables.

### Basic EFA

```{r efa-basic, eval = FALSE}
result <- perform_efa(
  mtcars,
  vars = c("mpg", "hp", "wt", "disp", "drat", "qsec"),
  n_factors = NULL, # Auto-determine
  rotation = "varimax",
  scores = TRUE
)

# Factor loadings
print(result$loadings)

# Variance explained
print(result$variance)

# Model fit
print(result$fit)

# Insights
cat(paste(result$insights, collapse = "\n"))
```

### Factor Retention

#### Parallel Analysis

Compare eigenvalues to random data:

```{r parallel-analysis, eval = FALSE}
result <- perform_efa(
  mtcars,
  n_factors = "parallel"
)

print(paste("Factors suggested by parallel analysis:", result$n_factors))
```

#### Scree Plot Criteria

Visual inspection of eigenvalue drop:

```{r scree-efa, eval = FALSE}
result <- perform_efa(
  mtcars,
  n_factors = "scree"
)
```

### Rotation Methods

#### Varimax (Orthogonal)

Assumes uncorrelated factors:

```{r varimax, eval = FALSE}
result <- perform_efa(
  mtcars,
  n_factors = 3,
  rotation = "varimax"
)
```

#### Promax (Oblique)

Allows correlated factors:

```{r promax, eval = FALSE}
result <- perform_efa(
  mtcars,
  n_factors = 3,
  rotation = "promax"
)

# Factor correlations
print(result$factor_correlations)
```

### Factor Interpretation

```{r efa-interpret, eval = FALSE}
result <- perform_efa(mtcars, n_factors = 2)

# Interpret factors based on high loadings
for (i in 1:result$n_factors) {
  cat(sprintf("\n--- Factor %d ---\n", i))

  # Variables with loadings > 0.4
  high_loadings <- result$loadings[abs(result$loadings[, i]) > 0.4, i]
  print(sort(high_loadings, decreasing = TRUE))
}
```

## Generalized Linear Models

Fit and compare GLMs with automatic family detection.

### Automatic Family Selection

```{r glm-auto, eval = FALSE}
# Auto-detect appropriate family based on outcome
result <- perform_glm(
  formula = mpg ~ hp + wt + cyl,
  data = mtcars,
  family = "auto",
  diagnostics = TRUE
)

# Model summary
print(result$summary)

# Diagnostics
print(result$diagnostics)

# Insights
cat(paste(result$insights, collapse = "\n"))
```

### Specific Families

#### Gaussian (Linear Regression)

```{r glm-gaussian, eval = FALSE}
result <- perform_glm(
  formula = mpg ~ hp + wt,
  data = mtcars,
  family = "gaussian"
)

print(result$coefficients)
```

#### Binomial (Logistic Regression)

```{r glm-binomial, eval = FALSE}
result <- perform_glm(
  formula = am ~ mpg + hp,
  data = mtcars,
  family = "binomial"
)

# Odds ratios
print(exp(result$coefficients$estimate))
```

#### Poisson (Count Data)

```{r glm-poisson, eval = FALSE}
result <- perform_glm(
  formula = cyl ~ mpg + hp,
  data = mtcars,
  family = "poisson"
)
```

### Model Diagnostics

Comprehensive diagnostic checks:

```{r glm-diagnostics, eval = FALSE}
result <- perform_glm(
  formula = mpg ~ hp + wt,
  data = mtcars,
  diagnostics = TRUE
)

# Residual diagnostics
print(result$diagnostics$residual_checks)

# Influence measures
print(result$diagnostics$influential_points)

# Model fit
print(result$diagnostics$goodness_of_fit)
```

### Model Comparison

Compare multiple models:

```{r model-comparison, eval = FALSE}
# Fit several models
model1 <- perform_glm(mpg ~ hp, data = mtcars)
model2 <- perform_glm(mpg ~ hp + wt, data = mtcars)
model3 <- perform_glm(mpg ~ hp + wt + cyl, data = mtcars)

# Compare using AIC
comparison <- compare_models(
  model1,
  model2,
  model3,
  criterion = "aic"
)

print(comparison)

# Best model
best_model <- comparison$models[[which.min(comparison$criterion_values)]]
```

## Non-Parametric Tests

Alternative tests when assumptions are violated.

### Mann-Whitney U Test

Non-parametric alternative to t-test:

```{r mann-whitney, eval = FALSE}
result <- perform_mann_whitney(
  x = mtcars$mpg[mtcars$am == 0],
  y = mtcars$mpg[mtcars$am == 1],
  alternative = "two.sided"
)

print(result)
# Includes rank-biserial correlation (effect size)
```

### Kruskal-Wallis Test

Non-parametric alternative to ANOVA:

```{r kruskal-wallis, eval = FALSE}
result <- perform_kruskal_wallis(
  formula = mpg ~ factor(cyl),
  data = mtcars,
  post_hoc = TRUE
)

# Overall test
print(result$test)

# Post-hoc comparisons (Dunn's test)
print(result$post_hoc)

# Effect size (epsilon-squared)
print(result$effect_size)
```

### Wilcoxon Signed-Rank Test

Paired non-parametric test:

```{r wilcoxon, eval = FALSE}
before <- c(120, 125, 130, 135, 128)
after <- c(115, 118, 125, 130, 122)

result <- perform_wilcoxon_signed(
  x = before,
  y = after,
  paired = TRUE
)

print(result)
```

### Friedman Test

Non-parametric repeated measures ANOVA:

```{r friedman, eval = FALSE}
# Requires data in long format with blocking variable
result <- perform_friedman(
  formula = value ~ condition | subject,
  data = my_repeated_measures_data
)

print(result)
# Includes Kendall's W (effect size)
```

## Summary

This vignette covered:

1. **Effect Sizes**: Cohen's d, Hedges' g, η², ω², Cramér's V, phi
2. **Normality Testing**: Multiple tests with transformation suggestions
3. **Missing Data**: Pattern analysis, MCAR testing, recommendations
4. **Outlier Detection**: Multiple methods with consensus approach
5. **PCA**: Component extraction, interpretation, rotation
6. **EFA**: Factor extraction, rotation, interpretation
7. **GLM**: Automatic family selection, diagnostics, comparison
8. **Non-Parametric Tests**: Rank-based alternatives

All methods include:
- Comprehensive output
- Effect sizes where applicable
- Narrative insights
- Publication-ready results

For visualization of these analyses, see the "Visualization and Reporting" vignette.
