---
title: "Getting Started with descriptR"
author: "Abdul Ali"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with descriptR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

**descriptR** is a comprehensive R package for descriptive and inferential statistics that automatically detects variable types and applies appropriate analyses. It generates narrative insights from data and produces publication-ready outputs in multiple formats (HTML, Word, Excel, PDF).

### Key Features

- **Intelligent Analysis**: Automatically detects variable types and selects appropriate statistical methods
- **Comprehensive Statistics**: From basic descriptives to advanced multivariate methods
- **Narrative Insights**: Generates human-readable interpretations of results
- **Multiple Export Formats**: HTML, Word, Excel, Markdown, and more
- **Publication-Ready**: Multiple journal-specific templates (APA, Nature, Science, NEJM)
- **Colorblind-Safe Palettes**: 16 carefully designed color schemes
- **Tidyverse Integration**: Works seamlessly with dplyr, ggplot2, and the tidyverse

## Installation

```{r installation, eval = FALSE}
# Install from GitHub
devtools::install_github("drabdulali/descriptR")

# Load the package
library(descriptR)
```

## Quick Start

### Basic Descriptive Statistics

The simplest way to get started is with the `analyze_and_report()` function, which performs analysis and generates a report in one step:

```{r quick-start, eval = FALSE}
library(descriptR)

# Load example data
data(mtcars)

# Quick descriptive analysis with HTML report
analyze_and_report(
  mtcars,
  output_file = "mtcars_report",
  analysis_type = "descriptive",
  format = "html",
  title = "Motor Trend Car Analysis"
)
```

This single function call:
- Analyzes all numeric variables in the dataset
- Generates descriptive statistics (mean, SD, quartiles, etc.)
- Creates narrative insights about the data
- Produces a professionally formatted HTML report

### Grouped Comparisons

Compare groups with automatic effect size calculations:

```{r grouped-comparison, eval = FALSE}
# Load iris dataset
data(iris)

# Compare species with grouped analysis
analyze_and_report(
  iris,
  output_file = "iris_comparison",
  analysis_type = "grouped",
  group = "Species",
  format = "html",
  template = "apa"
)
```

This analysis:
- Compares all numeric variables across species
- Calculates means, standard deviations by group
- Computes effect sizes for group differences
- Generates insights about notable differences
- Formats output in APA style

## Core Analysis Types

### 1. Descriptive Statistics

```{r descriptive, eval = FALSE}
# Descriptive analysis for specific variables
result <- perform_descriptive_analysis(
  mtcars,
  vars = c("mpg", "hp", "wt")
)

# View results
print(result$statistics)
print(result$insights)
```

**What you get:**
- Sample size and missing data counts
- Central tendency (mean, median)
- Dispersion (SD, IQR, range)
- Quartiles and extremes
- Narrative insights about variability and missing data

### 2. Correlation Analysis

```{r correlation, eval = FALSE}
# Correlation analysis
result <- perform_correlation_analysis(
  mtcars,
  vars = c("mpg", "hp", "wt", "disp"),
  method = "pearson"
)

# View correlation matrix
print(result$correlation_matrix)

# View insights
print(result$insights)
```

**What you get:**
- Correlation matrix
- P-values for each correlation
- Identification of strong correlations
- Insights about relationships

### 3. Normality Assessment

```{r normality, eval = FALSE}
# Assess normality
result <- perform_normality_analysis(
  mtcars,
  vars = c("mpg", "hp")
)

# View test results
print(result$normality_tests)

# Insights about distribution
print(result$insights)
```

**What you get:**
- Shapiro-Wilk test results
- Statistical significance
- Binary normality assessment
- Recommendations for parametric vs non-parametric methods

### 4. Missing Data Analysis

```{r missing, eval = FALSE}
# Create data with missing values
data_with_na <- mtcars
data_with_na$mpg[1:5] <- NA
data_with_na$hp[3:7] <- NA

# Analyze missing patterns
result <- perform_missing_analysis(data_with_na)

# View missing summary
print(result$missing_summary)
print(result$insights)
```

**What you get:**
- Missing counts and percentages per variable
- Identification of variables with high missingness
- Recommendations for handling missing data

### 5. Outlier Detection

```{r outliers, eval = FALSE}
# Detect outliers
result <- perform_outlier_analysis(
  mtcars,
  vars = c("mpg", "hp"),
  method = "zscore"
)

# View outlier summary
print(result$outlier_summary)
print(result$insights)
```

**What you get:**
- Outlier counts per variable
- Percentage of outliers
- Method used for detection
- Insights about data quality

### 6. Principal Component Analysis

```{r pca, eval = FALSE}
# Perform PCA
result <- perform_pca_analysis(
  mtcars,
  vars = c("mpg", "hp", "wt", "disp"),
  scale = TRUE
)

# View variance explained
print(result$variance_explained)

# View loadings
print(result$loadings)

# Insights
print(result$insights)
```

**What you get:**
- Variance explained by each component
- Cumulative variance
- Component loadings
- Component scores
- Insights about dimensionality

### 7. Comprehensive Analysis

Run all applicable analyses in one call:

```{r comprehensive, eval = FALSE}
# Comprehensive analysis
result <- perform_comprehensive_analysis(
  mtcars,
  vars = c("mpg", "hp", "wt"),
  group = "cyl"
)

# Access individual analyses
print(result$analyses$descriptive)
print(result$analyses$normality)
print(result$analyses$correlation)

# View compiled insights
print(result$insights)
```

**What you get:**
- Descriptive statistics
- Missing data analysis
- Normality assessment
- Outlier detection
- Correlation analysis
- Grouped comparisons (if group specified)
- All insights compiled in one place

## Exporting Results

### Multiple Format Export

Export your analysis to multiple formats simultaneously:

```{r export-formats, eval = FALSE}
# Analyze data
result <- perform_descriptive_analysis(mtcars)

# Export to all formats
generate_report(
  result,
  output_file = "mtcars_analysis",
  format = "all", # Exports HTML, Word, Excel, and Markdown
  template = "apa",
  title = "Motor Trend Car Analysis",
  author = "Your Name"
)
```

### Format-Specific Export

Export to a single format:

```{r export-single, eval = FALSE}
# HTML only
generate_report(
  result,
  output_file = "report.html",
  format = "html",
  template = "nature"
)

# Word only
generate_report(
  result,
  output_file = "report.docx",
  format = "word",
  template = "science"
)

# Excel only
generate_report(
  result,
  output_file = "report.xlsx",
  format = "excel"
)
```

### Available Templates

Choose from journal-specific templates:

- **default**: Clean, modern design
- **apa**: American Psychological Association style
- **nature**: Nature journal style
- **science**: Science journal style
- **nejm**: New England Journal of Medicine style

```{r templates, eval = FALSE}
# Use Nature template
analyze_and_report(
  mtcars,
  "nature_style_report",
  analysis_type = "descriptive",
  format = "html",
  template = "nature"
)
```

## Working with Results

### Accessing Result Components

All analysis functions return structured objects:

```{r result-structure, eval = FALSE}
result <- perform_descriptive_analysis(mtcars)

# Main statistics table
stats <- result$statistics
print(stats)

# Narrative insights
insights <- result$insights
cat(paste(insights, collapse = "\n"))

# Metadata
meta <- result$metadata
print(meta$timestamp)
print(meta$n_variables)
```

### Extracting Specific Statistics

```{r extract-stats, eval = FALSE}
# Get mean for specific variable
mpg_mean <- result$statistics[result$statistics$variable == "mpg", "mean"]

# Get all means
all_means <- result$statistics[, c("variable", "mean")]

# Filter variables with high variability
high_var <- result$statistics[result$statistics$sd > 10, ]
```

## Quick Data Export

### Export Data to Excel

Use `quick_export_excel()` for fast data export:

```{r quick-excel, eval = FALSE}
# Simple export
quick_export_excel(
  mtcars,
  "mtcars_data.xlsx",
  sheet_name = "Car Data"
)
```

### Batch Export Tables

Export multiple data frames to a single Excel workbook:

```{r batch-export, eval = FALSE}
# Create multiple analysis results
desc <- perform_descriptive_analysis(mtcars)
corr <- perform_correlation_analysis(mtcars)

# Export to Excel with multiple sheets
batch_export_tables(
  list(
    descriptive = desc$statistics,
    correlations = corr$correlations
  ),
  output_file = "analysis_results.xlsx",
  format = "excel"
)
```

## Best Practices

### 1. Always Inspect Your Data First

```{r best-practices-1, eval = FALSE}
# Check structure
str(mtcars)

# Summary
summary(mtcars)

# Missing data
result <- perform_missing_analysis(mtcars)
print(result$missing_summary)
```

### 2. Choose Appropriate Analysis Type

```{r best-practices-2, eval = FALSE}
# Check normality before parametric tests
normality <- perform_normality_analysis(mtcars, vars = "mpg")

if (normality$normality_tests$normal[1]) {
  # Use parametric methods
  message("Data is normal - use parametric tests")
} else {
  # Use non-parametric methods
  message("Data is non-normal - use non-parametric tests")
}
```

### 3. Check for Outliers

```{r best-practices-3, eval = FALSE}
# Detect outliers before analysis
outliers <- perform_outlier_analysis(mtcars)

if (any(outliers$outlier_summary$n_outliers > 0)) {
  message("Outliers detected - consider sensitivity analysis")
}
```

### 4. Document Your Analysis

```{r best-practices-4, eval = FALSE}
# Use comprehensive analysis for complete documentation
result <- analyze_and_report(
  mtcars,
  "complete_analysis",
  analysis_type = "comprehensive",
  format = "all", # Create documentation in all formats
  title = "Complete Motor Trend Analysis",
  author = "Your Name",
  include_plots = TRUE
)
```

## Next Steps

- Learn about **advanced statistical methods** in the "Advanced Analysis" vignette
- Explore **visualization and color schemes** in the "Visualization and Reporting" vignette
- Check the function reference for detailed parameter documentation

## Getting Help

```{r help, eval = FALSE}
# Function help
?analyze_and_report
?generate_report

# Package help
help(package = "descriptR")

# Report issues
# https://github.com/drabdulali/descriptR/issues
```

## Summary

descriptR provides:

1. **One-function workflow**: `analyze_and_report()` does everything
2. **Modular analysis**: Individual functions for specific analyses
3. **Multiple export formats**: HTML, Word, Excel, Markdown
4. **Professional templates**: APA, Nature, Science, NEJM
5. **Automatic insights**: Narrative interpretation of results
6. **Publication-ready**: High-quality outputs for papers and presentations

Start simple with `analyze_and_report()`, then explore individual functions as needed!
