---
title: "Visualization and Reporting with descriptR"
author: "Abdul Ali"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization and Reporting with descriptR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

**descriptR** provides comprehensive visualization and reporting capabilities, including:

- Automatic plot type selection
- 16 colorblind-safe color schemes
- Diagnostic visualizations
- Multi-format export (PNG, PDF, SVG, etc.)
- Journal-specific report templates
- Publication-ready outputs

```{r load, eval = FALSE}
library(descriptR)
library(ggplot2)
```

## Color Schemes

descriptR includes 16 carefully designed, colorblind-safe color palettes.

### Available Schemes

Four base schemes with four boldness levels each:

- **subtle**: Muted, understated colors (publications)
- **medium**: Balanced visibility (default)
- **bold**: High saturation (presentations)
- **vivid**: Maximum impact (posters, social media)

Each scheme has 4 boldness levels (1-4), creating 16 total combinations.

### Basic Usage

```{r colors-basic, eval = FALSE}
# Get default color scheme
colors <- get_color_scheme(
  scheme = "medium",
  boldness = 2,
  n = 8
)

print(colors)
```

### Scheme Selection

```{r colors-schemes, eval = FALSE}
# Subtle colors for publications
subtle <- get_color_scheme("subtle", boldness = 2, n = 5)

# Bold colors for presentations
bold <- get_color_scheme("bold", boldness = 3, n = 5)

# Vivid colors for posters
vivid <- get_color_scheme("vivid", boldness = 4, n = 5)

# Preview colors
preview_color_scheme("medium", boldness = 2, n = 8)
```

### Number of Colors

Request any number of colors (palette extends automatically):

```{r colors-n, eval = FALSE}
# Small palette
colors_3 <- get_color_scheme("medium", n = 3)

# Large palette (interpolates additional colors)
colors_20 <- get_color_scheme("medium", n = 20)
```

### Reverse Palette

Reverse color order:

```{r colors-reverse, eval = FALSE}
colors_normal <- get_color_scheme("medium", n = 5, reverse = FALSE)
colors_reversed <- get_color_scheme("medium", n = 5, reverse = TRUE)
```

### Scheme Recommendations

Get recommended scheme for specific contexts:

```{r colors-recommend, eval = FALSE}
# Publication
pub_rec <- get_scheme_for_type("categorical", "publication")
pub_colors <- get_color_scheme(pub_rec$scheme, pub_rec$boldness)

# Presentation
pres_rec <- get_scheme_for_type("categorical", "presentation")
pres_colors <- get_color_scheme(pres_rec$scheme, pres_rec$boldness)

# Web
web_rec <- get_scheme_for_type("continuous", "web")
web_colors <- get_color_scheme(web_rec$scheme, web_rec$boldness)
```

## Specialized Palettes

### Diverging Palettes

For data with a meaningful midpoint (e.g., correlation matrices):

```{r diverging, eval = FALSE}
# Create diverging palette
div_colors <- create_diverging_palette(
  low_color = "#0571B0",
  mid_color = "#F7F7F7",
  high_color = "#CA0020",
  n = 11
)

# Use with correlation heatmap
plot_correlation_matrix(
  mtcars,
  color_palette = div_colors
)
```

### Sequential Palettes

For ordered data (light to dark or vice versa):

```{r sequential, eval = FALSE}
# Light to dark
seq_colors <- create_sequential_palette(
  base_color = "#3182BD",
  n = 9,
  direction = "light_to_dark"
)

# Dark to light
seq_colors_rev <- create_sequential_palette(
  base_color = "#3182BD",
  n = 9,
  direction = "dark_to_light"
)
```

## Color Accessibility

Check WCAG compliance for color combinations.

### Contrast Checking

```{r accessibility, eval = FALSE}
# Check contrast ratio
result <- check_color_accessibility(
  foreground = "#000000",
  background = "#FFFFFF",
  level = "AA"
)

print(result$contrast_ratio)
print(result$passes_normal) # Normal text
print(result$passes_large)  # Large text (18pt+)
```

### AAA Standard

For maximum accessibility:

```{r accessibility-aaa, eval = FALSE}
result <- check_color_accessibility(
  foreground = "#333333",
  background = "#FFFFFF",
  level = "AAA"
)

if (result$passes_normal) {
  message("Passes AAA for normal text")
}
```

## Exporting Colors

Export palettes in various formats.

### Hex Format

```{r export-hex, eval = FALSE}
colors <- get_color_scheme("medium", n = 5)
hex_colors <- export_color_palette(colors, format = "hex")
print(hex_colors)
```

### RGB Format

```{r export-rgb, eval = FALSE}
rgb_colors <- export_color_palette(colors, format = "rgb")
# Returns: "rgb(65, 105, 225)", "rgb(220, 20, 60)", ...
```

### CSS Variables

For web development:

```{r export-css, eval = FALSE}
css_vars <- export_color_palette(
  colors,
  format = "css",
  name = "my_palette"
)

# Returns CSS custom properties:
# :root {
#   --my-palette-1: #4169E1;
#   --my-palette-2: #DC143C;
#   ...
# }
```

### R Code

For sharing with colleagues:

```{r export-r, eval = FALSE}
r_code <- export_color_palette(
  colors,
  format = "r",
  name = "my_colors"
)

# Returns:
# my_colors <- c("#4169E1", "#DC143C", ...)
```

## Automatic Plotting

The `plot_variable()` function automatically selects appropriate plot types.

### Univariate Plots

#### Continuous Variables

Automatically creates histogram or density plot:

```{r plot-continuous, eval = FALSE}
# Auto-select plot type
plot <- plot_variable(
  mtcars,
  x = "mpg",
  plot_type = "auto",
  color_scheme = "medium",
  theme = "publication"
)

print(plot)
```

#### Categorical Variables

Creates bar chart:

```{r plot-categorical, eval = FALSE}
plot <- plot_variable(
  mtcars,
  x = "cyl",
  plot_type = "auto"
)
```

### Bivariate Plots

#### Continuous vs Continuous

Creates scatter plot with optional smoothing:

```{r plot-scatter, eval = FALSE}
plot <- plot_variable(
  mtcars,
  x = "wt",
  y = "mpg",
  plot_type = "auto",
  add_smooth = TRUE
)
```

#### Continuous vs Categorical

Creates boxplot or violin plot:

```{r plot-boxplot, eval = FALSE}
# Boxplot
plot <- plot_variable(
  mtcars,
  x = "cyl",
  y = "mpg",
  plot_type = "boxplot"
)

# Violin plot
plot <- plot_variable(
  mtcars,
  x = "cyl",
  y = "mpg",
  plot_type = "violin"
)
```

### Grouped Plots

Add color grouping:

```{r plot-grouped, eval = FALSE}
plot <- plot_variable(
  iris,
  x = "Sepal.Length",
  y = "Sepal.Width",
  group = "Species",
  plot_type = "scatter",
  color_scheme = "bold"
)
```

### Plot Types

Override auto-detection:

```{r plot-types, eval = FALSE}
# Available types:
# - "histogram"
# - "density"
# - "boxplot"
# - "violin"
# - "scatter"
# - "bar"
# - "qq"

plot <- plot_variable(
  mtcars,
  x = "mpg",
  plot_type = "density",
  fill_color = "#3182BD"
)
```

## Diagnostic Plots

Specialized visualizations for analysis results.

### Normality Diagnostics

Combined diagnostic panel:

```{r plot-normality, eval = FALSE}
# Creates 4-panel diagnostic:
# - Histogram with normal curve
# - Q-Q plot
# - Density plot
# - Box plot

plot <- plot_normality_diagnostics(
  mtcars,
  variable = "mpg"
)

print(plot)
```

### Scree Plots

For PCA/EFA:

```{r plot-scree, eval = FALSE}
# Perform PCA
pca_result <- perform_pca(mtcars)

# Create scree plot
plot <- plot_scree(
  pca_result,
  add_kaiser = TRUE # Add Kaiser criterion line
)

print(plot)
```

### Missing Data Patterns

Visualize missing data structure:

```{r plot-missing, eval = FALSE}
# Create data with missing values
data_na <- mtcars
data_na$mpg[1:5] <- NA
data_na$hp[3:8] <- NA

# Analyze missing data
missing_result <- analyze_missing(data_na)

# Visualize patterns
plot <- plot_missing_pattern(
  data_na,
  max_patterns = 20
)

print(plot)
```

### Outlier Visualization

Show outliers in context:

```{r plot-outliers, eval = FALSE}
# Detect outliers
outlier_result <- detect_outliers(
  mtcars,
  vars = "mpg",
  methods = "zscore"
)

# Visualize
plot <- plot_outliers(
  outlier_result,
  data = mtcars,
  variable = "mpg"
)

print(plot)
```

### Effect Size Plots

Forest plots for effect sizes:

```{r plot-effects, eval = FALSE}
# Example effect sizes with confidence intervals
effects <- c(0.3, 0.5, 0.8)
ci_lower <- c(0.1, 0.3, 0.6)
ci_upper <- c(0.5, 0.7, 1.0)
labels <- c("Effect 1", "Effect 2", "Effect 3")

plot <- plot_effect_sizes(
  effects = effects,
  ci_lower = ci_lower,
  ci_upper = ci_upper,
  labels = labels,
  effect_type = "Cohen's d"
)

print(plot)
```

### Group Comparisons

Visualize group differences:

```{r plot-group-comparison, eval = FALSE}
plot <- plot_group_comparison(
  iris,
  outcome = "Sepal.Length",
  group = "Species",
  plot_type = "violin", # or "boxplot", "density"
  show_points = TRUE,
  color_scheme = "bold"
)

print(plot)
```

## Correlation Matrices

Heatmaps with hierarchical clustering:

```{r plot-correlation, eval = FALSE}
# Create correlation matrix plot
plot <- plot_correlation_matrix(
  mtcars,
  vars = c("mpg", "hp", "wt", "disp", "drat"),
  method = "pearson",
  cluster = TRUE,
  show_values = TRUE,
  color_palette = NULL # Uses diverging palette
)

print(plot)
```

### Customization

```{r correlation-custom, eval = FALSE}
# Custom diverging palette
my_palette <- create_diverging_palette(
  low_color = "#0571B0",
  mid_color = "#FFFFFF",
  high_color = "#CA0020",
  n = 11
)

plot <- plot_correlation_matrix(
  mtcars,
  color_palette = my_palette,
  cluster = FALSE,
  show_values = TRUE
)
```

## Themes

Pre-configured themes for different contexts.

### Publication Theme

Clean, minimal theme for journals:

```{r theme-publication, eval = FALSE}
plot <- plot_variable(
  mtcars,
  x = "mpg",
  theme = "publication"
)

# Equivalent to:
plot + theme_publication()
```

### Presentation Theme

Larger fonts and elements:

```{r theme-presentation, eval = FALSE}
plot <- plot_variable(
  mtcars,
  x = "mpg",
  theme = "presentation"
)
```

### Poster Theme

Maximum visibility:

```{r theme-poster, eval = FALSE}
plot <- plot_variable(
  mtcars,
  x = "mpg",
  theme = "poster"
)
```

## Saving Plots

Export plots with multiple formats and resolutions.

### Single Plot Export

```{r save-plot, eval = FALSE}
plot <- plot_variable(mtcars, x = "mpg")

# Save as PNG (300 DPI)
save_plot(
  plot,
  filename = "mpg_histogram.png",
  width = 7,
  height = 5,
  dpi = 300
)

# Save as PDF (vector)
save_plot(
  plot,
  filename = "mpg_histogram.pdf",
  width = 7,
  height = 5
)

# Save as SVG (vector)
save_plot(
  plot,
  filename = "mpg_histogram.svg",
  width = 7,
  height = 5
)
```

### High-Resolution Export

For publications and posters:

```{r save-high-res, eval = FALSE}
# Publication quality (600 DPI)
save_plot(plot, "figure1_600dpi.png", dpi = 600)

# Ultra high-res for posters (1200 DPI)
save_plot(plot, "poster_figure.png", dpi = 1200)
```

### Batch Export

Export multiple plots at once:

```{r export-plots-batch, eval = FALSE}
# Create several plots
plots <- list(
  histogram = plot_variable(mtcars, x = "mpg"),
  scatter = plot_variable(mtcars, x = "wt", y = "mpg"),
  boxplot = plot_variable(mtcars, x = "cyl", y = "mpg", plot_type = "boxplot")
)

# Export all as PNG and PDF
export_plots(
  plots,
  output_dir = "figures",
  prefix = "mtcars",
  formats = c("png", "pdf"),
  dpi = 300
)

# Creates:
# figures/mtcars_histogram.png
# figures/mtcars_histogram.pdf
# figures/mtcars_scatter.png
# figures/mtcars_scatter.pdf
# figures/mtcars_boxplot.png
# figures/mtcars_boxplot.pdf
```

## Report Generation

Create comprehensive reports with multiple output formats.

### HTML Reports

```{r report-html, eval = FALSE}
# Analyze data
result <- perform_descriptive_analysis(mtcars)

# Generate HTML report
generate_report(
  result,
  output_file = "analysis_report.html",
  format = "html",
  template = "default",
  title = "Motor Trend Car Analysis",
  author = "Your Name",
  include_plots = TRUE
)
```

### Journal Templates

Use journal-specific formatting:

```{r report-templates, eval = FALSE}
# APA style
generate_report(
  result,
  output_file = "apa_report.html",
  format = "html",
  template = "apa"
)

# Nature style
generate_report(
  result,
  output_file = "nature_report.html",
  template = "nature"
)

# Science style
generate_report(
  result,
  output_file = "science_report.html",
  template = "science"
)

# NEJM style
generate_report(
  result,
  output_file = "nejm_report.html",
  template = "nejm"
)
```

### Word Documents

For collaborative editing:

```{r report-word, eval = FALSE}
generate_report(
  result,
  output_file = "report.docx",
  format = "word",
  title = "Analysis Report",
  author = "Your Name"
)
```

### Excel Workbooks

Multi-sheet workbooks with formatted tables:

```{r report-excel, eval = FALSE}
generate_report(
  result,
  output_file = "results.xlsx",
  format = "excel"
)

# Creates workbook with sheets:
# - Metadata
# - Statistics
# - Insights (if available)
```

### Markdown

Plain text format:

```{r report-markdown, eval = FALSE}
generate_report(
  result,
  output_file = "report.md",
  format = "markdown"
)
```

### Multiple Formats

Export to all formats at once:

```{r report-all, eval = FALSE}
generate_report(
  result,
  output_file = "comprehensive_report", # No extension
  format = "all", # Creates .html, .docx, .xlsx, .md
  template = "apa",
  title = "Comprehensive Analysis",
  author = "Your Name"
)
```

## One-Step Workflow

Analyze and report in a single function call:

```{r workflow-complete, eval = FALSE}
# Complete workflow: analyze + report
analyze_and_report(
  data = mtcars,
  output_file = "mtcars_analysis",
  analysis_type = "comprehensive",
  format = "all",
  template = "apa",
  title = "Comprehensive Motor Trend Analysis",
  author = "Your Name",
  include_plots = TRUE
)

# This creates:
# - mtcars_analysis.html (APA formatted)
# - mtcars_analysis.docx (Word document)
# - mtcars_analysis.xlsx (Excel workbook)
# - mtcars_analysis.md (Markdown file)
# All with plots, tables, and narrative insights
```

## Custom Report Components

### Adding Custom Content

```{r custom-report, eval = FALSE}
# Create analysis
result <- perform_descriptive_analysis(mtcars)

# Add custom insights
result$insights <- c(
  result$insights,
  "Custom insight: Cars with higher MPG tend to be lighter",
  "Custom insight: Horsepower shows bimodal distribution"
)

# Add custom metadata
result$metadata$custom_field <- "Custom value"
result$metadata$analysis_date <- Sys.Date()

# Generate report with custom content
generate_report(
  result,
  output_file = "custom_report.html",
  format = "html",
  title = "Custom Analysis"
)
```

### Creating Custom Tables

```{r custom-tables, eval = FALSE}
# Create custom summary tables
tables <- list(
  descriptive = perform_descriptive_analysis(mtcars)$statistics,
  correlation = perform_correlation_analysis(mtcars)$correlations,
  normality = perform_normality_analysis(mtcars)$normality_tests
)

# Export as Excel
batch_export_tables(
  tables,
  output_file = "analysis_tables.xlsx",
  format = "excel"
)

# Or as CSV
batch_export_tables(
  tables,
  output_file = "analysis_tables", # Directory
  format = "csv"
)
```

## Best Practices

### 1. Color Selection

```{r best-colors, eval = FALSE}
# Publications: subtle, low boldness
pub_colors <- get_color_scheme("subtle", boldness = 2)

# Presentations: bold, medium-high boldness
pres_colors <- get_color_scheme("bold", boldness = 3)

# Posters: vivid, high boldness
poster_colors <- get_color_scheme("vivid", boldness = 4)

# Always check accessibility
check_color_accessibility(
  foreground = pub_colors[1],
  background = "#FFFFFF"
)
```

### 2. Plot Export

```{r best-export, eval = FALSE}
# Publications: 300-600 DPI, PDF/PNG
save_plot(plot, "figure1.pdf", width = 7, height = 5)
save_plot(plot, "figure1.png", width = 7, height = 5, dpi = 600)

# Presentations: 150-300 DPI, PNG
save_plot(plot, "slide_figure.png", dpi = 150)

# Posters: 300-600 DPI, large size
save_plot(plot, "poster.png", width = 20, height = 15, dpi = 300)
```

### 3. Report Formatting

```{r best-reports, eval = FALSE}
# Match template to target venue
generate_report(result, "journal_submission.docx",
                format = "word", template = "apa")

# Include all details in comprehensive format
generate_report(result, "full_analysis",
                format = "all", include_plots = TRUE)

# Share data with Excel
generate_report(result, "data_tables.xlsx", format = "excel")
```

### 4. Reproducibility

```{r best-reproducibility, eval = FALSE}
# Document your analysis completely
result <- analyze_and_report(
  data = mtcars,
  output_file = "reproducible_analysis",
  analysis_type = "comprehensive",
  format = "all",
  template = "default",
  title = paste("Analysis performed on", Sys.Date()),
  author = "Your Name",
  include_plots = TRUE
)

# Save session info
writeLines(
  capture.output(sessionInfo()),
  "session_info.txt"
)
```

## Summary

descriptR visualization and reporting features:

1. **16 colorblind-safe palettes** across 4 schemes × 4 boldness levels
2. **Automatic plot type selection** based on variable types
3. **Diagnostic visualizations** for all analysis types
4. **Multiple export formats**: PNG, PDF, SVG, TIFF, EPS
5. **DPI control**: 72 (screen) to 1200+ (ultra high-res)
6. **5 journal templates**: Default, APA, Nature, Science, NEJM
7. **4 output formats**: HTML, Word, Excel, Markdown
8. **One-step workflow**: `analyze_and_report()` does everything

All visualizations are publication-ready with minimal customization required!
